---
title: "How Can Observations of Clinical Trial Termination Risk Factors Help Researchers Identify and Address Potential Problems Before Trials Are Terminated?"
subtitle: "We can utilize observations across thousands of clinical drug trials, including those that have failed, to predict if a trial may terminate. Using those insights, how might we be able to prevent trial failure?"
about:
  template: jolla
  links:
    - text: Termination Signals
      href: termination-signals.qmd
      icon: activity
    - text: What Termination Looks Like
      href: trial-termination.qmd
      icon: eye
    - text: Prevention & Benefits
      href: prevention.qmd
      icon: shield-check
---

## Where studies stand

<div class="status-graphic">
<img src="assets/aact_status_distribution.svg" alt="Horizontal bar chart showing all 14 overall-status categories among 602,514 AACT studies. Terminated studies are highlighted in coral, with counts and percentages shown for every category." />
</div>

This foundational view captures all **602,514** registered clinical studies in the Aggregate Analysis of ClinicalTrials.gov (AACT) candidate dataset, maintaining each reported status category independently.

A central finding immediately emerges: **34,295 studies were explicitly marked `TERMINATED`** (representing **5.7%** of all registered trials, and **9.4%** of closed completed or terminated trials). Rather than treating trial termination as a rare anomaly, the data reveals that premature study termination is a recurring challenge across the clinical research enterprise.

::: {.callout-note}
### Key Takeaway from Status Distribution
While completed studies represent the largest single group (329,096 trials), over 34,000 trials halted prematurely. Understanding the operational, design, and clinical factors associated with these terminations provides a blueprint for early detection and prevention.
:::

## Key Variables & Data Dimensions Gathered

To evaluate termination risk factors comprehensively, we gathered and synthesized multidimensional indicators across 602,514 studies:

<div class="metrics-container">
  <div class="metric-card">
    <div class="metric-number">602,514</div>
    <div class="metric-label">Total Studies</div>
    <div class="metric-sub">Complete AACT spine analyzed</div>
  </div>
  <div class="metric-card">
    <div class="metric-number">34,295</div>
    <div class="metric-label">Terminated Trials</div>
    <div class="metric-sub">5.7% of total; 9.4% of resolved trials</div>
  </div>
  <div class="metric-card alt">
    <div class="metric-number">459,840</div>
    <div class="metric-label">Interventional Trials</div>
    <div class="metric-sub">vs 140,623 observational studies</div>
  </div>
  <div class="metric-card alt">
    <div class="metric-number">30,742</div>
    <div class="metric-label">Reported Reasons</div>
    <div class="metric-sub">Documented causes of premature halts</div>
  </div>
</div>

### Variables Collected Across Clinical Dimensions

| Domain | Key Variables Gathered | Description & Analytical Value |
|:---|:---|:---|
| **Study Identification & Status** | `nct_id`, `overall_status`, `terminated_binary`, `why_stopped` | Tracks trial status across 14 categories and captures free-text stopping rationales for over 30,000 terminated studies. |
| **Recruitment & Scale** | `enrollment`, `enrollment_type` | Quantifies planned and actual participant cohorts, distinguishing between anticipated targets and actual accrual. |
| **Development Phase** | `phase` (Phase I, II, III, IV, Mixed, NA) | Evaluates how failure rates evolve from early safety testing to large-scale confirmatory trials. |
| **Sponsorship & Funding** | `lead_sponsor_type`, `lead_sponsor_organization`, `sponsor_count` | Categorizes sponsors into Industry, Academic/Hospital, NIH, and Federal agencies to assess funding stability. |
| **Study Architecture & Complexity** | `randomized_vs_nonrandomized`, `placebo_used`, `number_of_arms`, `number_of_sites` | Captures structural protocol complexity, including blinding, control arms, and global facility footprints. |
| **Timeline & Duration** | `start_year`, `study_start_date`, `study_completion_date`, `study_duration_days` | Tracks temporal evolution over 36 years (1990–2026) and measures elapsed trial duration before resolution. |

## Therapeutic Areas & Disease Distribution

Clinical trial volume and termination vulnerability vary widely across medical disciplines. The statistics chart below maps registered trials and termination rates across major therapeutic areas:

```{r}
#| label: fig-therapeutic-areas
#| fig-cap: "Clinical trial volume and termination rate across primary therapeutic areas"
#| fig-width: 12.5
#| fig-height: 7.2
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(readr)
library(scales)

df_area <- read_tsv("data/processed/summary_therapeutic_areas.tsv", show_col_types = FALSE) |>
  filter(therapeutic_area != "Unspecified / Missing") |>
  mutate(therapeutic_area = factor(therapeutic_area, levels = rev(therapeutic_area)))

ggplot(df_area, aes(x = total_studies, y = therapeutic_area)) +
  geom_col(fill = "#b8d8d8", width = 0.68) +
  geom_col(aes(x = terminated_studies), fill = "#e76f51", width = 0.68) +
  geom_text(
    aes(label = paste0(comma(total_studies), " studies (", percent(termination_rate, accuracy = 0.1), " terminated)")),
    hjust = -0.04,
    color = "#183642",
    size = 3.3
  ) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.28))) +
  labs(
    x = "Total Registered Studies",
    y = NULL,
    subtitle = "Mint bars indicate total registered studies; coral bars denote studies marked TERMINATED."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", color = "#183642", size = 9.8),
    axis.text.x = element_text(color = "#42616a"),
    axis.title.x = element_text(color = "#183642", face = "bold", margin = margin(t = 8)),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```

::: {.callout-note}
### Therapeutic Area Insights
- **Oncology Dominance & Risk**: Oncology / Cancer represents the largest identified clinical indication with **110,360 studies (18.3% of all trials)**, and simultaneously exhibits the highest termination rate at **9.3%** (over 10,200 terminated trials). Oncology trials face severe patient accrual hurdles due to narrow biomarker inclusion criteria and toxicities.
- **Healthy Volunteer Baseline**: Studies on healthy volunteers and normal controls display the lowest termination rate (**3.0%**), underscoring that patient disease severity and complex eligibility directly elevate termination hazard.
:::

## Are trials getting larger? Trial enrollment over time

A critical question when predicting trial feasibility is whether clinical trials have grown in scale over the decades. The two coordinated visualizations below combine the **longitudinal trajectory of planned enrollment** with an **interactive panel of histograms** illustrating how trial sizes are distributed across eras.

### 1. Longitudinal Enrollment Trajectory (1990–2026)

This view tracks median planned enrollment (solid coral line) and the middle 50% interquartile band (shaded mint band) on a logarithmic scale for 575,628 trials with positive reported enrollment.

```{r}
#| label: fig-home-enrollment
#| fig-cap: "The typical and middle 50% of reported trial enrollment by study start year (1990–2026)"
#| fig-width: 12.5
#| fig-height: 5.5
#| echo: false
#| message: false
#| warning: false

annual_enrollment <- read_tsv(
  "data/processed/summary_annual_enrollment.tsv",
  show_col_types = FALSE
)

ggplot(annual_enrollment, aes(x = start_year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#b8d8d8", alpha = 0.85) +
  geom_line(aes(y = median), color = "#e76f51", linewidth = 1.15) +
  geom_point(aes(y = median), color = "#183642", fill = "#e76f51", shape = 21, size = 2.4, stroke = 0.5) +
  scale_y_log10(labels = comma, breaks = c(10, 100, 1000, 10000, 100000, 1000000)) +
  scale_x_continuous(breaks = seq(1990, 2025, by = 5), expand = expansion(mult = c(0.01, 0.02))) +
  labs(
    x = "Study start year",
    y = "Reported enrollment (log scale)",
    subtitle = "Solid coral line represents yearly median; shaded band shows the interquartile range (25th to 75th percentiles)."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "#183642"),
    axis.title = element_text(color = "#183642", face = "bold"),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```

### 2. Interactive Panel of Histograms: Enrollment Distribution by Era

The histogram panel below details the distribution of participant enrollment across four distinct historical eras: **1990–1999**, **2000–2009**, **2010–2019**, and **2020–2026**. Hover over any bar to inspect the exact share and number of trials within that enrollment bin, or click eras in the legend to compare subsets.

```{r}
#| label: fig-home-histograms
#| echo: false
#| message: false
#| warning: false

library(plotly)

hist_data <- read_tsv(
  "data/processed/summary_enrollment_histograms.tsv",
  show_col_types = FALSE
)

hist_data$bin_label <- factor(hist_data$bin_label, levels = unique(hist_data$bin_label))

era_colors <- c(
  "1990–1999" = "#42616a",
  "2000–2009" = "#b8d8d8",
  "2010–2019" = "#183642",
  "2020–2026" = "#e76f51"
)

p_hist <- plot_ly()

for (era_name in unique(hist_data$era)) {
  sub_era <- filter(hist_data, era == era_name)
  p_hist <- add_bars(
    p_hist,
    data = sub_era,
    x = ~bin_label,
    y = ~pct,
    name = era_name,
    marker = list(
      color = era_colors[era_name],
      line = list(color = "#ffffff", width = 0.5)
    ),
    hovertemplate = paste0(
      "<b>", era_name, " Era</b><br>",
      "Cohort Range: %{x} participants<br>",
      "Share of Trials: %{y:.1f}%<br>",
      "Total Trials in Bin: ", comma(sub_era$count),
      "<extra></extra>"
    )
  )
}

p_hist <- layout(
  p_hist,
  barmode = "group",
  bargap = 0.15,
  bargroupgap = 0.05,
  font = list(family = "DM Sans", color = "#183642"),
  paper_bgcolor = "#f7f4ed",
  plot_bgcolor = "#f7f4ed",
  xaxis = list(
    title = list(text = "Planned Enrollment Range", font = list(size = 12, color = "#183642")),
    tickangle = -35,
    tickfont = list(size = 10, color = "#183642"),
    gridcolor = "rgba(24, 54, 66, 0.08)"
  ),
  yaxis = list(
    title = list(text = "Share of Trials in Era (%)", font = list(size = 12, color = "#183642")),
    ticksuffix = "%",
    tickfont = list(size = 10, color = "#183642"),
    gridcolor = "rgba(24, 54, 66, 0.1)"
  ),
  legend = list(
    title = list(text = "<b>Study Era</b>"),
    orientation = "h",
    x = 0.15,
    y = 1.12
  ),
  margin = list(t = 50, b = 80, l = 60, r = 30)
)

p_hist
```

::: {.callout-note}
### Key Insights from Enrollment Distribution
1. **Convergence on Modern Target Sizes**: In the 1990s (prior to mandated registry reporting), recorded trials were disproportionately large landmark trials. Since the mid-2000s, clinical research has settled into a consistent distribution, with the bulk of trials enrolling between **25 and 250 participants** (typical Phase I–II studies).
2. **The Long Tail of Mega-Trials**: While median trial size has remained relatively stable (~60–80 patients), roughly 10% of trials plan for 1,000+ participants, introducing dramatic operational complexity and multi-site coordination risks.
:::

## Important Note

Correlation does **NOT** equal causation. Instead, data related to the risk factors for the termination of drug trials are merely *associated* with drug trial failure. Identifying these statistical associations allows research teams to flag operational vulnerabilities and implement risk-mitigation strategies early in trial planning.