---
title: "How Can Observations of Clinical Trial Termination Risk Factors Help Researchers Identify and Address Potential Problems Before Trials Are Terminated?"
subtitle: "We can utilize observations across thousands of clinical drug trials, including those that have failed, to predict if a trial may terminate. Using those insights, how might we be able to prevent trial failure?"
about:
  template: jolla
  links:
    - text: Termination Signals
      href: termination-signals.qmd
      icon: activity
    - text: What Termination Looks Like
      href: trial-termination.qmd
      icon: eye
    - text: Prevention & Benefits
      href: prevention.qmd
      icon: shield-check
---

## Where studies stand

<div class="status-graphic">
<img src="assets/aact_status_distribution.svg" alt="Horizontal bar chart showing all 14 overall-status categories among 602,514 AACT studies. Terminated studies are highlighted in coral, with counts and percentages shown for every category." />
</div>

This foundational view captures all **602,514** registered clinical studies in the Aggregate Analysis of ClinicalTrials.gov (AACT) candidate dataset, maintaining each reported status category independently.

A central finding immediately emerges: **34,295 studies were explicitly marked `TERMINATED`** (representing **5.7%** of all registered trials, and **9.4%** of closed completed or terminated trials). Rather than treating trial termination as a rare anomaly, the data reveals that premature study termination is a recurring challenge across the clinical research enterprise.

::: {.callout-note}
### Key Takeaway from Status Distribution
While completed studies represent the largest single group (329,096 trials), over 34,000 trials halted prematurely. Understanding the operational, design, and clinical factors associated with these terminations provides a blueprint for early detection and prevention.
:::

## Therapeutic Areas & Disease Distribution

Clinical trial volume and termination vulnerability vary widely across medical disciplines. The statistics chart below maps registered trials and termination rates across major therapeutic areas:

```{r}
#| label: home-therapeutic-areas
#| fig-width: 12.5
#| fig-height: 7.2
#| echo: false
#| message: false
#| warning: false

library(plotly)
library(htmlwidgets)
library(scales)
library(dplyr)
library(readr)
library(scales)

df_area <- read_tsv("data/processed/summary_therapeutic_areas.tsv", show_col_types = FALSE) |>
  filter(therapeutic_area != "Unspecified / Missing") |>
  mutate(therapeutic_area = factor(therapeutic_area, levels = rev(therapeutic_area)))

ggplot(df_area, aes(x = total_studies, y = therapeutic_area)) +
  geom_col(fill = "#b8d8d8", width = 0.68) +
  geom_col(aes(x = terminated_studies), fill = "#e76f51", width = 0.68) +
  geom_text(
    aes(label = paste0(comma(total_studies), " studies (", percent(termination_rate, accuracy = 0.1), " terminated)")),
    hjust = -0.04,
    color = "#183642",
    size = 3.3
  ) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.28))) +
  labs(
    x = "Total Registered Studies",
    y = NULL,
    subtitle = "Mint bars indicate total registered studies; coral bars denote studies marked TERMINATED."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", color = "#183642", size = 9.8),
    axis.text.x = element_text(color = "#42616a"),
    axis.title.x = element_text(color = "#183642", face = "bold", margin = margin(t = 8)),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```

::: {.callout-note}
### Therapeutic Area Insights
- **Oncology Dominance & Risk**: Oncology / Cancer represents the largest identified clinical indication with **110,360 studies (18.3% of all trials)**, and simultaneously exhibits the highest termination rate at **9.3%** (over 10,200 terminated trials). Oncology trials face severe patient accrual hurdles due to narrow biomarker inclusion criteria and toxicities.
- **Healthy Volunteer Baseline**: Studies on healthy volunteers and normal controls display the lowest termination rate (**3.0%**), underscoring that patient disease severity and complex eligibility directly elevate termination hazard.
:::

## What do individual trials look like over time?

Summary bars can hide the studies they count. This view shows the individual records directly: each point is one study plotted by its start year and planned enrollment. Coral points are studies marked `TERMINATED`; teal points are all other status categories. The logarithmic y-axis keeps small and very large studies visible together.

```{r}
#| label: home-enrollment-trend
#| echo: false
#| message: false
#| warning: false

library(plotly)
library(dplyr)
library(readr)

study_points <- read_tsv(
  "data/processed/aact_trial_termination_candidate_2026-09-12.tsv.gz",
  show_col_types = FALSE,
  progress = FALSE
) |>
  filter(!is.na(start_year), !is.na(enrollment), enrollment > 0) |>
  mutate(
    termination_label = if_else(terminated_binary == 1, "Terminated", "Not terminated"),
    termination_label = factor(termination_label, levels = c("Not terminated", "Terminated"))
  )

set.seed(42)
not_terminated_count <- sum(study_points$terminated_binary == 0)
not_terminated_points <- study_points |>
  filter(terminated_binary == 0) |>
  slice_sample(n = min(18000L, not_terminated_count))

terminated_points <- study_points |>
  filter(terminated_binary == 1)

plot_points <- bind_rows(not_terminated_points, terminated_points) |>
  mutate(
    hover_label = paste0("NCT: ", nct_id),
    terminated_binary = as.integer(terminated_binary)
  )

BRAND_CORAL <- "#e76f51"
BRAND_NAVY  <- "#183642"
BRAND_TEAL  <- "#42616a"
BRAND_BG    <- "#f7f4ed"
BRAND_FONT  <- "DM Sans, system-ui, -apple-system, sans-serif"

plot_ly(
  data = plot_points,
  height = 700,
  source = "study-enrollment-scatter"
) |>
  add_trace(
    data = filter(plot_points, terminated_binary == 0),
    x = ~start_year,
    y = ~enrollment,
    type = "scatter",
    mode = "markers",
    name = "Not terminated",
    marker = list(
      color = BRAND_TEAL,
      opacity = 0.22,
      size = 5,
      line = list(color = "rgba(24, 54, 66, 0.15)", width = 0.6)
    ),
    customdata = ~hover_label,
    hovertemplate = paste0(
      "<b>%{customdata}</b><br>",
      "Start year: <b>%{x}</b><br>",
      "Planned enrollment: <b>%{y:,}</b><br>",
      "Status: <b>Not terminated</b><extra></extra>"
    )
  ) |>
  add_trace(
    data = filter(plot_points, terminated_binary == 1),
    x = ~start_year,
    y = ~enrollment,
    type = "scatter",
    mode = "markers",
    name = "Terminated",
    marker = list(
      color = BRAND_CORAL,
      opacity = 0.85,
      size = 6,
      line = list(color = "rgba(24, 54, 66, 0.35)", width = 0.8)
    ),
    customdata = ~hover_label,
    hovertemplate = paste0(
      "<b>%{customdata}</b><br>",
      "Start year: <b>%{x}</b><br>",
      "Planned enrollment: <b>%{y:,}</b><br>",
      "Status: <b>Terminated</b><extra></extra>"
    )
  ) |>
  layout(
    dragmode = "zoom",
    hovermode = "closest",
    paper_bgcolor = BRAND_BG,
    plot_bgcolor = BRAND_BG,
    font = list(family = BRAND_FONT, color = BRAND_NAVY),
    xaxis = list(
      title = list(text = "Study Start Year", font = list(size = 14, color = BRAND_NAVY)),
      range = c(1990, 2026),
      tickvals = seq(1990, 2025, by = 5),
      gridcolor = "rgba(24, 54, 66, 0.08)",
      tickfont = list(size = 12, color = BRAND_NAVY),
      zeroline = FALSE
    ),
    yaxis = list(
      title = list(text = "Planned Enrollment per Study (log scale)", font = list(size = 14, color = BRAND_NAVY)),
      type = "log",
      tickvals = c(1, 10, 100, 1000, 10000, 100000, 1000000),
      ticktext = c("1", "10", "100", "1k", "10k", "100k", "1M"),
      gridcolor = "rgba(24, 54, 66, 0.08)",
      tickfont = list(size = 12, color = BRAND_NAVY),
      zeroline = FALSE
    ),
    legend = list(
      orientation = "h",
      x = 0.02,
      y = 1.02,
      yanchor = "bottom",
      font = list(size = 12, color = BRAND_NAVY)
    ),
    margin = list(t = 35, b = 65, l = 85, r = 35),
    annotations = list(
      list(
        x = 0.0,
        y = 1.02,
        xref = "paper",
        yref = "paper",
        text = paste0("<b>", format(nrow(study_points), big.mark = ","), " studies with positive reported enrollment</b>"),
        showarrow = FALSE,
        font = list(size = 12, color = BRAND_TEAL)
      )
    )
  ) |>
  config(
    displaylogo = FALSE,
    responsive = TRUE,
    scrollZoom = TRUE,
    modeBarButtonsToRemove = c("select2d", "lasso2d"),
    toImageButtonOptions = list(format = "png", filename = "study_enrollment_scatter")
  )
```

### 2. Interactive Panel of Histograms: Enrollment Distribution by Era

The histogram panel below details the distribution of participant enrollment across four distinct historical eras: **1990–1999**, **2000–2009**, **2010–2019**, and **2020–2026**. Use the toggle buttons above the chart to view each era individually with clean, wide bars and direct percentage labels, or select "Compare All" to view the side-by-side comparison.

```{r}
#| label: home-histograms-panel
#| echo: false
#| message: false
#| warning: false

library(plotly)

hist_data <- read_tsv(
  "data/processed/summary_enrollment_histograms.tsv",
  show_col_types = FALSE
)

hist_data$bin_label <- factor(hist_data$bin_label, levels = unique(hist_data$bin_label))
eras <- unique(hist_data$era)

era_colors <- c(
  "1990–1999" = "#42616a",
  "2000–2009" = "#b8d8d8",
  "2010–2019" = "#183642",
  "2020–2026" = "#e76f51"
)

BRAND_CORAL <- "#e76f51"
BRAND_NAVY  <- "#183642"
BRAND_TEAL  <- "#42616a"
BRAND_MINT  <- "#b8d8d8"
BRAND_BG    <- "#f7f4ed"
BRAND_FONT  <- "DM Sans, system-ui, -apple-system, sans-serif"

p_hist <- plot_ly(height = 500)

# Default to latest era (2020–2026) active
active_era_idx <- length(eras)

for (i in seq_along(eras)) {
  e <- eras[i]
  sub_data <- filter(hist_data, era == e) |>
    mutate(label_text = paste0(round(pct, 1), "%"))
  is_active <- (i == active_era_idx)
  
  p_hist <- add_bars(
    p_hist,
    data = sub_data,
    x = ~bin_label,
    y = ~pct,
    name = e,
    visible = is_active,
    marker = list(
      color = era_colors[e],
      line = list(color = BRAND_NAVY, width = 0.8)
    ),
    text = ~label_text,
    textposition = "outside",
    cliponaxis = FALSE,
    textfont = list(family = BRAND_FONT, size = 10, color = BRAND_NAVY),
    hovertemplate = paste0(
      "<b>", e, " Era</b><br>",
      "Cohort: <b>%{x}</b> participants<br>",
      "Share of Era: <b>%{y:.1f}%</b><br>",
      "Total Studies: <b>", comma(sub_data$count), "</b> of ", comma(sub_data$total_in_era[1]),
      "<extra></extra>"
    )
  )
}

# Build toggle buttons for each individual era + compare all
buttons <- list()
for (i in seq_along(eras)) {
  vis <- rep(FALSE, length(eras))
  vis[i] <- TRUE
  e <- eras[i]
  tot <- format(filter(hist_data, era == e)$total_in_era[1], big.mark = ",")
  
  buttons[[i]] <- list(
    method = "update",
    args = list(
      list(visible = vis),
      list(
        title = list(
          text = paste0("<b>Enrollment Distribution: ", e, " Era</b> (", tot, " studies)"),
          font = list(family = BRAND_FONT, size = 15, color = BRAND_NAVY),
          x = 0.0,
          y = 0.98,
          xanchor = "left",
          yanchor = "top"
        ),
        barmode = "group",
        showlegend = FALSE,
        "yaxis.range" = list(0, 26)
      )
    ),
    label = e
  )
}

# Add "Compare All" button
buttons[[length(buttons) + 1]] <- list(
  method = "update",
  args = list(
    list(visible = rep(TRUE, length(eras))),
    list(
      title = list(
        text = "<b>Enrollment Distribution: All Eras Compared Side-by-Side</b>",
        font = list(family = BRAND_FONT, size = 15, color = BRAND_NAVY),
        x = 0.0,
        y = 0.98,
        xanchor = "left",
        yanchor = "top"
      ),
      barmode = "group",
      showlegend = TRUE,
      "yaxis.range" = list(0, 26)
    )
  ),
  label = "Compare All"
)

active_e <- eras[active_era_idx]
active_tot <- format(filter(hist_data, era == active_e)$total_in_era[1], big.mark = ",")

p_hist <- layout(
  p_hist,
  title = list(
    text = paste0("<b>Enrollment Distribution: ", active_e, " Era</b> (", active_tot, " studies)"),
    font = list(family = BRAND_FONT, size = 15, color = BRAND_NAVY),
    x = 0.0,
    y = 0.98,
    xanchor = "left",
    yanchor = "top"
  ),
  barmode = "group",
  font = list(family = BRAND_FONT, color = BRAND_NAVY),
  paper_bgcolor = BRAND_BG,
  plot_bgcolor = BRAND_BG,
  hoverlabel = list(bgcolor = BRAND_NAVY, font = list(color = "#ffffff", family = BRAND_FONT)),
  xaxis = list(
    title = list(text = "Planned Enrollment Cohort Range", font = list(size = 12, color = BRAND_NAVY)),
    tickangle = -35,
    tickfont = list(size = 10, color = BRAND_NAVY),
    gridcolor = "rgba(24, 54, 66, 0.08)"
  ),
  yaxis = list(
    title = list(text = "Share of Trials in Era (%)", font = list(size = 12, color = BRAND_NAVY)),
    ticksuffix = "%",
    range = c(0, 26),
    tickfont = list(size = 10, color = BRAND_NAVY),
    gridcolor = "rgba(24, 54, 66, 0.1)"
  ),
  updatemenus = list(list(
    type = "buttons",
    direction = "right",
    x = 0.0,
    y = 1.12,
    xanchor = "left",
    yanchor = "top",
    active = active_era_idx - 1,
    buttons = buttons,
    font = list(family = BRAND_FONT, size = 11, color = BRAND_NAVY)
  )),
  legend = list(
    title = list(text = "<b>Study Era</b>"),
    orientation = "h",
    x = 0.50,
    y = 1.12,
    yanchor = "top"
  ),
  margin = list(t = 110, b = 75, l = 60, r = 30)
) |>
  config(displaylogo = FALSE, responsive = TRUE)

p_hist
```

::: {.callout-note}
### Key Insights from Enrollment Distribution
1. **Convergence on Modern Target Sizes**: In the 1990s (prior to mandated registry reporting), recorded trials were disproportionately large landmark trials. Since the mid-2000s, clinical research has settled into a consistent distribution, with the bulk of trials enrolling between **25 and 250 participants** (typical Phase I–II studies).
2. **The Long Tail of Mega-Trials**: While median trial size has remained relatively stable (~60–80 patients), roughly 10% of trials plan for 1,000+ participants, introducing dramatic operational complexity and multi-site coordination risks.
:::

## Important Note

Correlation does **NOT** equal causation. Instead, data related to the risk factors for the termination of drug trials are merely *associated* with drug trial failure. Identifying these statistical associations allows research teams to flag operational vulnerabilities and implement risk-mitigation strategies early in trial planning.