---
title: "Variables to Consider and Possible Links to Failures in Clinical Trials"
subtitle: "Empirical risk factors, complexity indicators, and early warning patterns observed across 602,514 studies."
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| message: false
library(ggplot2)
library(dplyr)
library(readr)
library(scales)
library(plotly)

# Load precomputed summaries
status_summary <- read_tsv("data/processed/summary_status.tsv", show_col_types = FALSE)
phase_summary <- read_tsv("data/processed/summary_phase.tsv", show_col_types = FALSE)
design_summary <- read_tsv("data/processed/summary_design.tsv", show_col_types = FALSE)
sponsor_summary <- read_tsv("data/processed/summary_sponsors.tsv", show_col_types = FALSE)
site_summary <- read_tsv("data/processed/summary_sites.tsv", show_col_types = FALSE)
hist_summary <- read_tsv("data/processed/summary_enrollment_histograms.tsv", show_col_types = FALSE)

# Signal summaries
df_interventions <- read_tsv("data/processed/summary_interventions.tsv", show_col_types = FALSE)
df_durations <- read_tsv("data/processed/summary_durations.tsv", show_col_types = FALSE)
df_randomization <- read_tsv("data/processed/summary_randomization.tsv", show_col_types = FALSE)
df_placebo <- read_tsv("data/processed/summary_placebo.tsv", show_col_types = FALSE)
df_complexity <- read_tsv("data/processed/summary_complexity.tsv", show_col_types = FALSE)
df_conditions <- read_tsv("data/processed/summary_conditions_count.tsv", show_col_types = FALSE)

# Unified Design Palette & Constants
BRAND_CORAL <- "#e76f51"
BRAND_NAVY  <- "#183642"
BRAND_TEAL  <- "#42616a"
BRAND_MINT  <- "#b8d8d8"
BRAND_AMBER <- "#f4a261"
BRAND_BG    <- "#f7f4ed"
BRAND_FONT  <- "DM Sans, system-ui, -apple-system, sans-serif"

# Master Helper for Interactive Category Charts
create_signal_chart <- function(
    df,
    cat_var,
    total_var,
    term_var,
    rate_var,
    x_label = "",
    y_label = "",
    highlight_cats = c(),
    is_horizontal = FALSE,
    reverse_cats = FALSE,
    annotation_text = NULL,
    annotation_cat = NULL,
    annotation_pos = NULL,
    custom_colors = NULL,
    insight_note = NULL,
    chart_height = 460
) {
  df_clean <- df |>
    mutate(
      cat_name = as.character(.data[[cat_var]]),
      total = .data[[total_var]],
      term = .data[[term_var]],
      rate = .data[[rate_var]],
      rate_pct = round(rate * 100, 1),
      other = total - term
    )
  
  if (!is.null(custom_colors)) {
    df_clean$bar_color <- unname(custom_colors)
  } else {
    df_clean$bar_color <- ifelse(df_clean$cat_name %in% highlight_cats, BRAND_CORAL, BRAND_TEAL)
  }
  
  if (reverse_cats) {
    df_clean$cat_name <- factor(df_clean$cat_name, levels = rev(df_clean$cat_name))
  } else {
    df_clean$cat_name <- factor(df_clean$cat_name, levels = df_clean$cat_name)
  }
  
  p <- plot_ly(data = df_clean, height = chart_height)
  
  if (!is_horizontal) {
    # VERTICAL BAR CHART
    p <- p |>
      add_bars(
        x = ~cat_name,
        y = ~rate_pct,
        name = "Termination Rate (%)",
        visible = TRUE,
        showlegend = FALSE,
        marker = list(
          color = ~bar_color,
          line = list(color = BRAND_NAVY, width = 0.8)
        ),
        hovertemplate = paste0(
          "<b>%{x}</b><br>",
          "Termination Rate: <b>%{y:.1f}%</b><br>",
          "Terminated Trials: <b>%{customdata[0]:,}</b><br>",
          "Total Trials: %{customdata[1]:,}<br>",
          "Completed / Active: %{customdata[2]:,}",
          if (!is.null(insight_note)) paste0("<br><i>", insight_note, "</i>") else "",
          "<extra></extra>"
        ),
        customdata = ~mapply(list, term, total, other, SIMPLIFY = FALSE)
      ) |>
      add_bars(
        x = ~cat_name,
        y = ~other,
        name = "Completed / Active",
        visible = FALSE,
        marker = list(color = BRAND_MINT, line = list(color = BRAND_TEAL, width = 0.8)),
        hovertemplate = paste0(
          "<b>%{x}</b><br>",
          "Completed / Active: <b>%{y:,}</b><br>",
          "Total Trials: %{customdata:,}<extra></extra>"
        ),
        customdata = ~total
      ) |>
      add_bars(
        x = ~cat_name,
        y = ~term,
        name = "Terminated",
        visible = FALSE,
        marker = list(color = BRAND_CORAL, line = list(color = BRAND_NAVY, width = 0.8)),
        hovertemplate = paste0(
          "<b>%{x}</b><br>",
          "Terminated Trials: <b>%{y:,}</b><br>",
          "Termination Rate: <b>", df_clean$rate_pct, "%</b><extra></extra>"
        )
      )
    
    annot_list <- list()
    if (!is.null(annotation_text) && !is.null(annotation_cat) && !is.null(annotation_pos)) {
      annot_list <- list(list(
        x = annotation_cat,
        y = annotation_pos,
        text = paste0("<b>", annotation_text, "</b>"),
        showarrow = TRUE,
        arrowhead = 2,
        arrowsize = 1,
        arrowwidth = 1.5,
        arrowcolor = BRAND_CORAL,
        ax = 0,
        ay = -40,
        font = list(family = BRAND_FONT, size = 11, color = BRAND_CORAL),
        bgcolor = "rgba(247, 244, 237, 0.95)",
        bordercolor = BRAND_CORAL,
        borderwidth = 1,
        borderpad = 4
      ))
    }
    
    p <- layout(
      p,
      barmode = "group",
      showlegend = FALSE,
      paper_bgcolor = BRAND_BG,
      plot_bgcolor = BRAND_BG,
      font = list(family = BRAND_FONT, color = BRAND_NAVY),
      hoverlabel = list(bgcolor = BRAND_NAVY, font = list(color = "#ffffff", family = BRAND_FONT)),
      xaxis = list(
        title = list(text = x_label, font = list(size = 12, color = BRAND_NAVY)),
        tickfont = list(size = 10, color = BRAND_NAVY),
        gridcolor = "rgba(24, 54, 66, 0.08)"
      ),
      yaxis = list(
        title = list(text = "Termination Rate (%)", font = list(size = 12, color = BRAND_NAVY)),
        ticksuffix = "%",
        gridcolor = "rgba(24, 54, 66, 0.1)"
      ),
      legend = list(orientation = "h", x = 0.45, y = 1.15, font = list(size = 11)),
      annotations = annot_list,
      updatemenus = list(list(
        type = "buttons",
        direction = "right",
        x = 0.0,
        y = 1.15,
        xanchor = "left",
        buttons = list(
          list(
            method = "update",
            args = list(
              list(visible = list(TRUE, FALSE, FALSE)),
              list(
                "yaxis.title.text" = "Termination Rate (%)",
                "yaxis.ticksuffix" = "%",
                "yaxis.tickformat" = "",
                "yaxis.autorange" = TRUE,
                barmode = "group",
                showlegend = FALSE,
                annotations = annot_list
              )
            ),
            label = "Termination Rate (%)"
          ),
          list(
            method = "update",
            args = list(
              list(visible = list(FALSE, TRUE, TRUE)),
              list(
                "yaxis.title.text" = "Number of Studies",
                "yaxis.ticksuffix" = "",
                "yaxis.tickformat" = ",d",
                "yaxis.autorange" = TRUE,
                barmode = "stack",
                showlegend = TRUE,
                annotations = list()
              )
            ),
            label = "Study Volume Breakdown"
          )
        ),
        font = list(family = BRAND_FONT, size = 11, color = BRAND_NAVY)
      )),
      margin = list(t = 65, b = 60, l = 60, r = 30)
    )
  } else {
    # HORIZONTAL BAR CHART
    p <- p |>
      add_bars(
        y = ~cat_name,
        x = ~rate_pct,
        name = "Termination Rate (%)",
        visible = TRUE,
        showlegend = FALSE,
        marker = list(
          color = ~bar_color,
          line = list(color = BRAND_NAVY, width = 0.8)
        ),
        hovertemplate = paste0(
          "<b>%{y}</b><br>",
          "Termination Rate: <b>%{x:.1f}%</b><br>",
          "Terminated Trials: <b>%{customdata[0]:,}</b><br>",
          "Total Trials: %{customdata[1]:,}<br>",
          "Completed / Active: %{customdata[2]:,}",
          if (!is.null(insight_note)) paste0("<br><i>", insight_note, "</i>") else "",
          "<extra></extra>"
        ),
        customdata = ~mapply(list, term, total, other, SIMPLIFY = FALSE)
      ) |>
      add_bars(
        y = ~cat_name,
        x = ~other,
        name = "Completed / Active",
        visible = FALSE,
        marker = list(color = BRAND_MINT, line = list(color = BRAND_TEAL, width = 0.8)),
        hovertemplate = paste0(
          "<b>%{y}</b><br>",
          "Completed / Active: <b>%{x:,}</b><br>",
          "Total Trials: %{customdata:,}<extra></extra>"
        ),
        customdata = ~total
      ) |>
      add_bars(
        y = ~cat_name,
        x = ~term,
        name = "Terminated",
        visible = FALSE,
        marker = list(color = BRAND_CORAL, line = list(color = BRAND_NAVY, width = 0.8)),
        hovertemplate = paste0(
          "<b>%{y}</b><br>",
          "Terminated Trials: <b>%{x:,}</b><br>",
          "Termination Rate: <b>", df_clean$rate_pct, "%</b><extra></extra>"
        )
      )
    
    annot_list <- list()
    if (!is.null(annotation_text) && !is.null(annotation_cat) && !is.null(annotation_pos)) {
      annot_list <- list(list(
        y = annotation_cat,
        x = annotation_pos,
        text = paste0("<b>", annotation_text, "</b>"),
        showarrow = TRUE,
        arrowhead = 2,
        arrowsize = 1,
        arrowwidth = 1.5,
        arrowcolor = BRAND_CORAL,
        ax = 45,
        ay = 0,
        font = list(family = BRAND_FONT, size = 11, color = BRAND_CORAL),
        bgcolor = "rgba(247, 244, 237, 0.95)",
        bordercolor = BRAND_CORAL,
        borderwidth = 1,
        borderpad = 4
      ))
    }
    
    p <- layout(
      p,
      barmode = "group",
      showlegend = FALSE,
      paper_bgcolor = BRAND_BG,
      plot_bgcolor = BRAND_BG,
      font = list(family = BRAND_FONT, color = BRAND_NAVY),
      hoverlabel = list(bgcolor = BRAND_NAVY, font = list(color = "#ffffff", family = BRAND_FONT)),
      yaxis = list(
        title = "",
        tickfont = list(size = 10, color = BRAND_NAVY)
      ),
      xaxis = list(
        title = list(text = "Termination Rate (%)", font = list(size = 12, color = BRAND_NAVY)),
        ticksuffix = "%",
        gridcolor = "rgba(24, 54, 66, 0.1)"
      ),
      legend = list(orientation = "h", x = 0.45, y = 1.15, font = list(size = 11)),
      annotations = annot_list,
      updatemenus = list(list(
        type = "buttons",
        direction = "right",
        x = 0.0,
        y = 1.15,
        xanchor = "left",
        buttons = list(
          list(
            method = "update",
            args = list(
              list(visible = list(TRUE, FALSE, FALSE)),
              list(
                "xaxis.title.text" = "Termination Rate (%)",
                "xaxis.ticksuffix" = "%",
                "xaxis.tickformat" = "",
                "xaxis.autorange" = TRUE,
                barmode = "group",
                showlegend = FALSE,
                annotations = annot_list
              )
            ),
            label = "Termination Rate (%)"
          ),
          list(
            method = "update",
            args = list(
              list(visible = list(FALSE, TRUE, TRUE)),
              list(
                "xaxis.title.text" = "Number of Studies",
                "xaxis.ticksuffix" = "",
                "xaxis.tickformat" = ",d",
                "xaxis.autorange" = TRUE,
                barmode = "stack",
                showlegend = TRUE,
                annotations = list()
              )
            ),
            label = "Study Volume Breakdown"
          )
        ),
        font = list(family = BRAND_FONT, size = 11, color = BRAND_NAVY)
      )),
      margin = list(t = 65, b = 60, l = 175, r = 30)
    )
  }
  
  p <- config(p, displaylogo = FALSE, responsive = TRUE)
  return(p)
}
```

## 1. Study Status & The Scope of Termination

Across the full registry of 602,514 studies, **34,295** are explicitly recorded as `TERMINATED`. The interactive chart below contrasts total registered studies against those marked terminated across all primary status categories. Use the buttons to toggle between linear and logarithmic scale to inspect low-frequency status categories, or hover over bars for exact counts and registry shares.

```{r}
#| label: signal-status

status_plot_data <- status_summary |>
  mutate(
    status_label = gsub("_", " ", overall_status),
    status_label = factor(status_label, levels = rev(gsub("_", " ", overall_status)))
  )

p_status <- plot_ly(data = status_plot_data, height = 560) |>
  add_bars(
    y = ~status_label,
    x = ~studies,
    name = "Total Registered Studies",
    marker = list(
      color = BRAND_MINT,
      line = list(color = BRAND_TEAL, width = 0.8)
    ),
    hovertemplate = paste0(
      "<b>%{y}</b><br>",
      "Total Registered: <b>%{x:,}</b><br>",
      "Share of Registry: <b>", status_plot_data$pct_of_total, "%</b><extra></extra>"
    )
  ) |>
  add_bars(
    y = ~status_label,
    x = ~terminated_count,
    name = "Terminated Subset",
    marker = list(
      color = BRAND_CORAL,
      line = list(color = BRAND_NAVY, width = 0.8)
    ),
    hovertemplate = paste0(
      "<b>%{y}</b><br>",
      "Terminated Studies: <b>%{x:,}</b><extra></extra>"
    )
  ) |>
  layout(
    barmode = "overlay",
    paper_bgcolor = BRAND_BG,
    plot_bgcolor = BRAND_BG,
    font = list(family = BRAND_FONT, color = BRAND_NAVY),
    hoverlabel = list(bgcolor = BRAND_NAVY, font = list(color = "#ffffff", family = BRAND_FONT)),
    xaxis = list(
      title = list(text = "Number of Studies (Linear Scale)", font = list(size = 12, color = BRAND_NAVY)),
      gridcolor = "rgba(24, 54, 66, 0.08)",
      tickformat = ",d"
    ),
    yaxis = list(title = "", tickfont = list(size = 10, color = BRAND_NAVY)),
    legend = list(orientation = "h", x = 0.42, y = 1.14, font = list(size = 11)),
    updatemenus = list(list(
      type = "buttons",
      direction = "right",
      x = 0.0,
      y = 1.14,
      xanchor = "left",
      buttons = list(
        list(
          method = "relayout",
          args = list(list(
            "xaxis.type" = "linear",
            "xaxis.title.text" = "Number of Studies (Linear Scale)"
          )),
          label = "Linear Scale"
        ),
        list(
          method = "relayout",
          args = list(list(
            "xaxis.type" = "log",
            "xaxis.title.text" = "Number of Studies (Log Scale - Reveals Rare Statuses)"
          )),
          label = "Log Scale"
        )
      ),
      font = list(family = BRAND_FONT, size = 11, color = BRAND_NAVY)
    )),
    margin = list(l = 180, r = 30, t = 60, b = 50)
  ) |>
  config(displaylogo = FALSE, responsive = TRUE)

p_status
```

::: {.callout-note}
### Understanding Closed vs Active Statuses
While the baseline termination rate across all registered records is 5.7%, among closed, definitive trials (completed + terminated: 363,391 studies), **the realized termination rate is 9.4%**—approaching 1 in every 10 trials.
:::

## 2. Trial Size Over Time: Interactive Panel of Histograms

Trial enrollment is one of the most critical structural constraints in clinical trial design. Under-recruitment directly leads to premature termination, while over-ambitious targets create severe operational drag.

The interactive histogram panel below reveals how planned participant cohorts are distributed across major research eras (**1990–1999**, **2000–2009**, **2010–2019**, and **2020–2026**). Use the dropdown menu in the upper-left of the chart to inspect specific eras or compare all eras side-by-side, or hover over bars to view exact trial counts and percentages.

```{r}
#| label: signal-interactive-histograms
#| echo: false
#| message: false
#| warning: false

hist_summary$bin_label <- factor(hist_summary$bin_label, levels = unique(hist_summary$bin_label))
eras <- unique(hist_summary$era)

era_colors <- c(
  "1990–1999" = "#42616a",
  "2000–2009" = "#b8d8d8",
  "2010–2019" = "#183642",
  "2020–2026" = "#e76f51"
)

p_panel <- plot_ly(height = 500)

for (i in seq_along(eras)) {
  e <- eras[i]
  sub_data <- filter(hist_summary, era == e)
  is_active <- (i == length(eras))
  
  p_panel <- add_bars(
    p_panel,
    data = sub_data,
    x = ~bin_label,
    y = ~count,
    name = e,
    visible = is_active,
    marker = list(
      color = era_colors[e],
      line = list(color = BRAND_NAVY, width = 0.6)
    ),
    hovertemplate = paste0(
      "<b>Era: ", e, "</b><br>",
      "Cohort: %{x} participants<br>",
      "Studies: %{y:,}<br>",
      "Share of Era: ", sub_data$pct, "%<extra></extra>"
    )
  )
}

buttons <- list()
for (i in seq_along(eras)) {
  vis <- rep(FALSE, length(eras))
  vis[i] <- TRUE
  buttons[[length(buttons) + 1]] <- list(
    method = "update",
    args = list(
      list(visible = vis),
      list(
        title = list(
          text = paste0("<b>Trial Enrollment Distribution (", eras[i], ")</b>"),
          font = list(family = BRAND_FONT, size = 15, color = BRAND_NAVY),
          x = 0.0,
          y = 0.98,
          xanchor = "left",
          yanchor = "top"
        ),
        barmode = "group",
        showlegend = FALSE
      )
    ),
    label = paste0("Era: ", eras[i])
  )
}

# Add "Compare All Eras" option
buttons[[length(buttons) + 1]] <- list(
  method = "update",
  args = list(
    list(visible = rep(TRUE, length(eras))),
    list(
      title = list(
        text = "<b>Trial Enrollment Distribution: All Eras Compared Side-by-Side</b>",
        font = list(family = BRAND_FONT, size = 15, color = BRAND_NAVY),
        x = 0.0,
        y = 0.98,
        xanchor = "left",
        yanchor = "top"
      ),
      barmode = "group",
      showlegend = TRUE
    )
  ),
  label = "Compare All Eras Side-by-Side"
)

p_panel <- layout(
  p_panel,
  showlegend = FALSE,
  title = list(
    text = paste0("<b>Trial Enrollment Distribution (", eras[length(eras)], ")</b>"),
    font = list(family = BRAND_FONT, size = 15, color = BRAND_NAVY),
    x = 0.0,
    y = 0.98,
    xanchor = "left",
    yanchor = "top"
  ),
  paper_bgcolor = BRAND_BG,
  plot_bgcolor = BRAND_BG,
  font = list(family = BRAND_FONT, color = BRAND_NAVY),
  hoverlabel = list(bgcolor = BRAND_NAVY, font = list(color = "#ffffff", family = BRAND_FONT)),
  xaxis = list(
    title = list(text = "Planned Enrollment Range", font = list(size = 11, color = BRAND_NAVY)),
    tickangle = -35,
    tickfont = list(size = 9, color = BRAND_NAVY),
    gridcolor = "rgba(24, 54, 66, 0.08)"
  ),
  yaxis = list(
    title = list(text = "Number of Trials", font = list(size = 11, color = BRAND_NAVY)),
    tickfont = list(size = 9, color = BRAND_NAVY),
    gridcolor = "rgba(24, 54, 66, 0.1)"
  ),
  updatemenus = list(list(
    type = "dropdown",
    active = length(eras) - 1,
    y = 1.12,
    x = 0.0,
    xanchor = "left",
    yanchor = "top",
    buttons = buttons,
    font = list(family = BRAND_FONT, size = 11, color = BRAND_NAVY)
  )),
  legend = list(
    orientation = "h",
    x = 0.40,
    y = 1.12,
    yanchor = "top",
    font = list(size = 10)
  ),
  margin = list(t = 110, b = 70, l = 60, r = 30)
) |>
  config(displaylogo = FALSE, responsive = TRUE)

p_panel
```

## 3. Trial Phase: The Phase II Vulnerability Spike

Phase of development is one of the strongest predictive signals for trial termination. As drugs transition from preliminary safety to formal proof-of-concept testing, the failure rate sharply peaks.

Use the interactive buttons above the chart to toggle between **Termination Rate (%)** and **Study Volume Breakdown**, or hover over bars to inspect exact counts and percentages.

```{r}
#| label: signal-phase

phase_plot_data <- phase_summary |>
  filter(!phase_display %in% c("Not applicable", "Other / Missing")) |>
  mutate(phase_display = factor(phase_display, levels = c("Phase I", "Mixed Phase (I/II or II/III)", "Phase II", "Phase III", "Phase IV"))) |>
  arrange(phase_display)

p_phase <- create_signal_chart(
  df = phase_plot_data,
  cat_var = "phase_display",
  total_var = "total_studies",
  term_var = "terminated_studies",
  rate_var = "termination_rate",
  x_label = "Development Phase",
  highlight_cats = c("Phase II", "Mixed Phase (I/II or II/III)"),
  annotation_text = "Peak: 11.1%<br>Valley of Death",
  annotation_cat = "Phase II",
  annotation_pos = 11.1,
  insight_note = "Phase II transitions to efficacy testing, creating the sharpest attrition peak.",
  chart_height = 470
)

p_phase
```

## 4. Study Architecture: Randomization & Placebo Use

The design of a trial's control mechanism directly impacts patient enrollment willingness, protocol compliance, and operational feasibility.

### A. Randomization Status

Non-randomized interventional studies experience a significantly higher termination rate (**8.4%**) than randomized trials (**5.5%**), often reflecting single-arm exploratory designs that are vulnerable to early discontinuation.

```{r}
#| label: signal-randomization

rand_plot_data <- df_randomization |>
  mutate(randomization_status = factor(randomization_status, levels = c("Randomized", "Non-randomized", "Not reported"))) |>
  arrange(randomization_status)

p_rand <- create_signal_chart(
  df = rand_plot_data,
  cat_var = "randomization_status",
  total_var = "total_studies",
  term_var = "terminated_studies",
  rate_var = "termination_rate",
  x_label = "Allocation / Randomization Architecture",
  highlight_cats = c("Non-randomized"),
  annotation_text = "8.4% Rate<br>+52% vs Randomized",
  annotation_cat = "Non-randomized",
  annotation_pos = 8.4,
  insight_note = "Single-arm exploratory designs are vulnerable to early discontinuation.",
  chart_height = 450
)

p_rand
```

### B. Placebo Use

Placebo-controlled trials demand rigorous double-blinding, identical manufacturing of active and sham agents, and ethical consent for potential inactive treatment. Consequently, trials with reported placebo controls exhibit an **8.6% termination rate**—more than 50% higher than trials without placebo controls (5.6%).

```{r}
#| label: signal-placebo

placebo_plot_data <- df_placebo |>
  mutate(placebo_status = factor(placebo_status, levels = c("Placebo control used", "No placebo reported", "Not reported"))) |>
  arrange(placebo_status)

p_placebo <- create_signal_chart(
  df = placebo_plot_data,
  cat_var = "placebo_status",
  total_var = "total_studies",
  term_var = "terminated_studies",
  rate_var = "termination_rate",
  x_label = "Placebo Reporting Status",
  highlight_cats = c("Placebo control used"),
  annotation_text = "8.6% Rate<br>+53% vs No Placebo",
  annotation_cat = "Placebo control used",
  annotation_pos = 8.6,
  insight_note = "Double-blind burdens and participant sham reluctance drive higher termination.",
  chart_height = 450
)

p_placebo
```

## 5. Protocol Scale: Interventions & Number of Conditions

### A. Number of Interventions

As trials incorporate multiple drugs, combination therapies, or complex dosing arms, the termination hazard rises steadily. The data reveals a clear dose-response relationship between intervention count and study termination:

```{r}
#| label: signal-interventions

int_plot_data <- df_interventions |>
  mutate(intervention_label = factor(intervention_label, levels = rev(intervention_label)))

p_int <- create_signal_chart(
  df = int_plot_data,
  cat_var = "intervention_label",
  total_var = "total_studies",
  term_var = "terminated_studies",
  rate_var = "termination_rate",
  x_label = "Termination Rate (%)",
  is_horizontal = TRUE,
  custom_colors = c(
    "0 (Observational/None)" = "#b8d8d8",
    "1 intervention"          = "#42616a",
    "2 interventions"         = "#42616a",
    "3 interventions"         = "#f4a261",
    "4 interventions"         = "#e76f51",
    "5+ interventions"        = "#e76f51"
  )[as.character(int_plot_data$intervention_label)],
  annotation_text = "8.2% (+53% vs Monotherapy)",
  annotation_cat = "5+ interventions",
  annotation_pos = 8.2,
  insight_note = "Multi-drug regimens compound toxicities, supply logistics, and dosing complexities.",
  chart_height = 470
)

p_int
```

### B. Number of Conditions Listed

Protocols targeting multiple comorbid conditions or broad diagnostic indications introduce diagnostic heterogeneity and patient subgroup stratification challenges:

```{r}
#| label: signal-conditions

cond_plot_data <- df_conditions |>
  filter(conditions_category != "0 / Missing") |>
  mutate(conditions_category = factor(conditions_category, levels = conditions_category))

p_cond <- create_signal_chart(
  df = cond_plot_data,
  cat_var = "conditions_category",
  total_var = "total_studies",
  term_var = "terminated_studies",
  rate_var = "termination_rate",
  x_label = "Number of Conditions Listed in Registry",
  highlight_cats = c("5+ conditions", "1 condition"),
  annotation_text = "6.0% Rate",
  annotation_cat = "5+ conditions",
  annotation_pos = 6.0,
  insight_note = "Broad indications introduce heterogeneous patient populations and strict stratification.",
  chart_height = 450
)

p_cond
```

## 6. Study Duration: The Attrition Timeline

How does elapsed study timeline correlate with termination? The chart below tracks termination rates across study duration brackets (from short studies under 6 months to multi-year longitudinal trials):

```{r}
#| label: signal-durations

dur_plot_data <- df_durations |>
  mutate(duration_bracket = factor(duration_bracket, levels = rev(duration_bracket)))

p_dur <- create_signal_chart(
  df = dur_plot_data,
  cat_var = "duration_bracket",
  total_var = "total_studies",
  term_var = "terminated_studies",
  rate_var = "termination_rate",
  x_label = "Termination Rate (%)",
  is_horizontal = TRUE,
  custom_colors = c(
    "< 6 months"  = "#b8d8d8",
    "6–12 months" = "#b8d8d8",
    "1–2 years"   = "#42616a",
    "2–3 years"   = "#e76f51",
    "3–5 years"   = "#e76f51",
    "5+ years"    = "#42616a"
  )[as.character(dur_plot_data$duration_bracket)],
  annotation_text = "Peak Danger: 6.8%",
  annotation_cat = "2–3 years",
  annotation_pos = 6.8,
  insight_note = "2–5 year window: trials endure accrual fatigue, PI turnover, and sponsor reprioritization.",
  chart_height = 470
)

p_dur
```

::: {.callout-note}
### The 2–5 Year Vulnerability Window
Short studies (< 6 months) resolve quickly before major institutional or competitive shifts occur (3.4% failure). Studies extending into years 2 through 5 endure prolonged exposure to funding renewals, principal investigator relocations, and patient dropouts.
:::

## 7. Protocol Complexity Score: Compounding Structural Risk

To capture the holistic operational burden of a clinical trial, we synthesized a **Composite Study Complexity Score (0 to 5 points)** incorporating:
- Multi-arm design ($\ge 3$ arms or $\ge 3$ interventions)
- Placebo control reporting
- Blinded masking (double, triple, or quadruple)
- Multi-site geographic footprint ($\ge 6$ facilities)
- Multi-endpoint burden ($\ge 3$ primary outcomes)

Stratifying 602,514 studies by complexity score reveals a stark, linear escalation in failure risk:

```{r}
#| label: signal-complexity

comp_plot_data <- df_complexity |>
  mutate(complexity_tier = factor(complexity_tier, levels = complexity_tier))

p_comp <- create_signal_chart(
  df = comp_plot_data,
  cat_var = "complexity_tier",
  total_var = "total_studies",
  term_var = "terminated_studies",
  rate_var = "termination_rate",
  x_label = "Protocol Complexity Tier",
  custom_colors = c(
    "Low Complexity (0–1)"       = "#b8d8d8",
    "Moderate Complexity (2)"     = "#42616a",
    "High Complexity (3)"         = "#f4a261",
    "Very High Complexity (4+)"   = "#e76f51"
  )[as.character(comp_plot_data$complexity_tier)],
  annotation_text = "9.3% Rate<br>+80% vs Low",
  annotation_cat = "Very High Complexity (4+)",
  annotation_pos = 9.3,
  insight_note = "Structural complexity nearly doubles termination rate from 5.2% to 9.3%.",
  chart_height = 470
)

p_comp
```

## 8. Sponsor Type & Site Network Footprint

### A. Lead Sponsor Category

```{r}
#| label: signal-sponsor

sponsor_plot_data <- sponsor_summary |>
  filter(total_studies >= 1000) |>
  arrange(desc(termination_rate)) |>
  mutate(sponsor_label = factor(sponsor_label, levels = rev(sponsor_label)))

p_sponsor <- create_signal_chart(
  df = sponsor_plot_data,
  cat_var = "sponsor_label",
  total_var = "total_studies",
  term_var = "terminated_studies",
  rate_var = "termination_rate",
  x_label = "Termination Rate (%)",
  is_horizontal = TRUE,
  highlight_cats = c("Industry (Pharma / Biotech)", "National Institutes of Health (NIH)"),
  annotation_text = "8.9% (Industry Leader)",
  annotation_cat = "Industry (Pharma / Biotech)",
  annotation_pos = 8.9,
  insight_note = "Industry actively terminates underperforming assets; NIH faces rigid grant windows.",
  chart_height = 480
)

p_sponsor
```

### B. Clinical Site Footprint

```{r}
#| label: signal-sites

site_plot_data <- site_summary |>
  filter(site_category != "0 sites / Not reported") |>
  mutate(site_category = factor(site_category, levels = c(
    "Single-site (1)",
    "Small multi-site (2–5)",
    "Medium multi-site (6–20)",
    "Large multi-site (21–50)",
    "Global mega-network (51+)"
  )))

p_site <- create_signal_chart(
  df = site_plot_data,
  cat_var = "site_category",
  total_var = "total_studies",
  term_var = "terminated_studies",
  rate_var = "termination_rate",
  x_label = "Operational Footprint (Number of Clinical Sites)",
  custom_colors = c(
    "Single-site (1)"          = "#b8d8d8",
    "Small multi-site (2–5)"   = "#42616a",
    "Medium multi-site (6–20)" = "#e76f51",
    "Large multi-site (21–50)" = "#e76f51",
    "Global mega-network (51+)"= "#e76f51"
  )[as.character(site_plot_data$site_category)],
  annotation_text = "10.3% Rate<br>2x Single-site",
  annotation_cat = "Medium multi-site (6–20)",
  annotation_pos = 10.3,
  insight_note = "Risk jumps from 5.0% to 10%+ once trials expand beyond a single institution.",
  chart_height = 470
)

p_site
```
