---
title: "Variables to Consider and Possible Links to Failures in Clinical Trials"
subtitle: "Empirical risk factors, complexity indicators, and early warning patterns observed across 602,514 studies."
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| message: false
library(ggplot2)
library(dplyr)
library(readr)
library(scales)
library(plotly)

# Load precomputed summaries
status_summary <- read_tsv("data/processed/summary_status.tsv", show_col_types = FALSE)
phase_summary <- read_tsv("data/processed/summary_phase.tsv", show_col_types = FALSE)
design_summary <- read_tsv("data/processed/summary_design.tsv", show_col_types = FALSE)
sponsor_summary <- read_tsv("data/processed/summary_sponsors.tsv", show_col_types = FALSE)
site_summary <- read_tsv("data/processed/summary_sites.tsv", show_col_types = FALSE)
hist_summary <- read_tsv("data/processed/summary_enrollment_histograms.tsv", show_col_types = FALSE)

# New signal summaries
df_interventions <- read_tsv("data/processed/summary_interventions.tsv", show_col_types = FALSE)
df_durations <- read_tsv("data/processed/summary_durations.tsv", show_col_types = FALSE)
df_randomization <- read_tsv("data/processed/summary_randomization.tsv", show_col_types = FALSE)
df_placebo <- read_tsv("data/processed/summary_placebo.tsv", show_col_types = FALSE)
df_complexity <- read_tsv("data/processed/summary_complexity.tsv", show_col_types = FALSE)
df_conditions <- read_tsv("data/processed/summary_conditions_count.tsv", show_col_types = FALSE)
```

## 1. Study Status & The Scope of Termination

Across the full registry of 602,514 studies, **34,295** are explicitly recorded as `TERMINATED`. The chart below contrasts total registered studies against those marked terminated across all primary status categories.

```{r}
#| label: fig-status
#| fig-cap: "Total registered studies and terminated subset by reported status category"
#| fig-width: 12.5
#| fig-height: 7.2

status_plot_data <- status_summary |>
  mutate(overall_status = factor(overall_status, levels = rev(overall_status)))

ggplot(status_plot_data, aes(x = studies, y = overall_status)) +
  geom_col(fill = "#b8d8d8", width = 0.68) +
  geom_col(aes(x = terminated_count), fill = "#e76f51", width = 0.68) +
  geom_text(aes(label = comma(studies)), hjust = -0.12, color = "#183642", size = 3.3) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.12))) +
  labs(
    x = "Number of studies",
    y = NULL,
    subtitle = "Mint bars indicate total registered studies; coral bars denote studies marked TERMINATED."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", color = "#183642", size = 10),
    axis.text.x = element_text(color = "#42616a"),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```

::: {.callout-note}
### Understanding Closed vs Active Statuses
While the baseline termination rate across all registered records is 5.7%, among closed, definitive trials (completed + terminated: 363,391 studies), **the realized termination rate is 9.4%**—approaching 1 in every 10 trials.
:::

## 2. Trial Size Over Time: Interactive Panel of Histograms

Trial enrollment is one of the most critical structural constraints in clinical trial design. Under-recruitment directly leads to premature termination, while over-ambitious targets create severe operational drag.

The interactive histogram panel below reveals how planned participant cohorts are distributed across major research eras (**1990–1999**, **2000–2009**, **2010–2019**, and **2020–2026**). Use the dropdown menu in the upper-left of the chart to inspect specific eras, or hover over bars to view exact trial counts and percentages.

```{r}
#| label: fig-interactive-histograms
#| echo: false
#| message: false
#| warning: false

hist_summary$bin_label <- factor(hist_summary$bin_label, levels = unique(hist_summary$bin_label))
eras <- unique(hist_summary$era)

p_panel <- plot_ly()

for (i in seq_along(eras)) {
  e <- eras[i]
  sub_data <- filter(hist_summary, era == e)
  is_active <- (i == length(eras))
  
  p_panel <- add_bars(
    p_panel,
    data = sub_data,
    x = ~bin_label,
    y = ~count,
    name = e,
    visible = is_active,
    marker = list(
      color = "#e76f51",
      line = list(color = "#183642", width = 0.6)
    ),
    hovertemplate = paste0(
      "<b>Era: ", e, "</b><br>",
      "Cohort: %{x} participants<br>",
      "Studies: %{y:,}<br>",
      "Share of Era: ", sub_data$pct, "%<extra></extra>"
    )
  )
}

buttons <- lapply(seq_along(eras), function(i) {
  vis <- rep(FALSE, length(eras))
  vis[i] <- TRUE
  list(
    method = "update",
    args = list(
      list(visible = vis),
      list(title = list(
        text = paste0("<b>Trial Enrollment Distribution (", eras[i], ")</b>"),
        font = list(family = "DM Sans", size = 15, color = "#183642")
      ))
    ),
    label = eras[i]
  )
})

p_panel <- layout(
  p_panel,
  title = list(
    text = paste0("<b>Trial Enrollment Distribution (", eras[length(eras)], ")</b>"),
    font = list(family = "DM Sans", size = 15, color = "#183642")
  ),
  xaxis = list(
    title = list(text = "Planned Enrollment Range", font = list(size = 11, color = "#183642")),
    tickangle = -35,
    tickfont = list(size = 9, color = "#183642"),
    gridcolor = "rgba(24, 54, 66, 0.08)"
  ),
  yaxis = list(
    title = list(text = "Number of Trials", font = list(size = 11, color = "#183642")),
    tickfont = list(size = 9, color = "#183642"),
    gridcolor = "rgba(24, 54, 66, 0.1)"
  ),
  paper_bgcolor = "#f7f4ed",
  plot_bgcolor = "#f7f4ed",
  updatemenus = list(list(
    type = "dropdown",
    active = length(eras) - 1,
    y = 1.18,
    x = 0.0,
    xanchor = "left",
    buttons = buttons,
    font = list(family = "DM Sans", size = 11, color = "#183642")
  )),
  margin = list(t = 60, b = 70, l = 60, r = 30)
)

p_panel
```

## 3. Trial Phase: The Phase II Vulnerability Spike

Phase of development is one of the strongest predictive signals for trial termination. As drugs transition from preliminary safety to formal proof-of-concept testing, the failure rate sharply peaks.

```{r}
#| label: fig-phase
#| fig-cap: "Trial termination rate by development phase"
#| fig-width: 12.5
#| fig-height: 5.8

phase_plot <- phase_summary |>
  filter(!phase_display %in% c("Not applicable", "Other / Missing")) |>
  mutate(phase_display = factor(phase_display, levels = c("Phase I", "Mixed Phase (I/II or II/III)", "Phase II", "Phase III", "Phase IV")))

ggplot(phase_plot, aes(x = phase_display, y = termination_rate)) +
  geom_col(fill = "#e76f51", width = 0.62) +
  geom_text(
    aes(label = paste0(percent(termination_rate, accuracy = 0.1), "\n(", comma(terminated_studies), " / ", comma(total_studies), ")")),
    vjust = -0.3,
    color = "#183642",
    size = 3.4,
    lineheight = 0.85
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.14), expand = c(0, 0)) +
  labs(
    x = "Development Phase",
    y = "Termination Rate",
    subtitle = "Phase II experiences the highest termination rate (11.1%), representing the classical 'valley of death' in drug development."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "#183642", size = 10),
    axis.title = element_text(color = "#183642", face = "bold"),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```

## 4. Study Architecture: Randomization & Placebo Use

The design of a trial's control mechanism directly impacts patient enrollment willingness, protocol compliance, and operational feasibility.

### A. Randomization Status

Non-randomized interventional studies experience a significantly higher termination rate (**8.4%**) than randomized trials (**5.5%**), often reflecting single-arm exploratory designs that are vulnerable to early discontinuation.

```{r}
#| label: fig-randomization
#| fig-cap: "Termination rate by study randomization status"
#| fig-width: 12.5
#| fig-height: 5.4

rand_plot <- df_randomization |>
  mutate(randomization_status = factor(randomization_status, levels = c("Randomized", "Non-randomized", "Not reported")))

ggplot(rand_plot, aes(x = randomization_status, y = termination_rate)) +
  geom_col(fill = "#e76f51", width = 0.55) +
  geom_text(
    aes(label = paste0(percent(termination_rate, accuracy = 0.1), "\n(", comma(terminated_studies), " / ", comma(total_studies), " trials)")),
    vjust = -0.25,
    color = "#183642",
    size = 3.4,
    lineheight = 0.9
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.11), expand = c(0, 0)) +
  labs(
    x = "Allocation / Randomization Architecture",
    y = "Termination Rate",
    subtitle = "Non-randomized studies encounter an 8.4% termination rate, compared to 5.5% in randomized designs."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "#183642", size = 10),
    axis.title = element_text(color = "#183642", face = "bold"),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```

### B. Placebo Use

Placebo-controlled trials demand rigorous double-blinding, identical manufacturing of active and sham agents, and ethical consent for potential inactive treatment. Consequently, trials with reported placebo controls exhibit an **8.6% termination rate**—more than 50% higher than trials without placebo controls (5.6%).

```{r}
#| label: fig-placebo
#| fig-cap: "Termination rate by placebo control reporting"
#| fig-width: 12.5
#| fig-height: 5.4

placebo_plot <- df_placebo |>
  mutate(placebo_status = factor(placebo_status, levels = c("Placebo control used", "No placebo reported", "Not reported")))

ggplot(placebo_plot, aes(x = placebo_status, y = termination_rate)) +
  geom_col(fill = "#42616a", width = 0.55) +
  geom_text(
    aes(label = paste0(percent(termination_rate, accuracy = 0.1), "\n(", comma(terminated_studies), " / ", comma(total_studies), " trials)")),
    vjust = -0.25,
    color = "#183642",
    size = 3.4,
    lineheight = 0.9
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.11), expand = c(0, 0)) +
  labs(
    x = "Placebo Reporting",
    y = "Termination Rate",
    subtitle = "Placebo-controlled studies terminate at 8.6%, driven by patient reluctance to receive sham therapy and double-blind logistical burdens."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "#183642", size = 10),
    axis.title = element_text(color = "#183642", face = "bold"),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```

## 5. Protocol Scale: Interventions & Number of Conditions

### A. Number of Interventions

As trials incorporate multiple drugs, combination therapies, or complex dosing arms, the termination hazard rises steadily. The data reveals a clear dose-response relationship between intervention count and study termination:

```{r}
#| label: fig-interventions
#| fig-cap: "Trial termination rate across number of tested interventions"
#| fig-width: 12.5
#| fig-height: 5.6

int_plot <- df_interventions |>
  mutate(intervention_label = factor(intervention_label, levels = rev(intervention_label)))

ggplot(int_plot, aes(x = termination_rate, y = intervention_label)) +
  geom_col(fill = "#e76f51", width = 0.65) +
  geom_text(
    aes(label = paste0(percent(termination_rate, accuracy = 0.1), " (", comma(terminated_studies), " / ", comma(total_studies), ")")),
    hjust = -0.06,
    color = "#183642",
    size = 3.3
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.11), expand = expansion(mult = c(0, 0.18))) +
  labs(
    x = "Termination Rate",
    y = NULL,
    subtitle = "Monotherapy trials fail at 5.3%, while complex regimens with 5+ interventions experience an 8.2% termination rate."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", color = "#183642", size = 9.5),
    axis.text.x = element_text(color = "#42616a"),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```

### B. Number of Conditions Listed

Protocols targeting multiple comorbid conditions or broad diagnostic indications introduce diagnostic heterogeneity and patient subgroup stratification challenges:

```{r}
#| label: fig-conditions
#| fig-cap: "Trial termination rate by number of reported conditions"
#| fig-width: 12.5
#| fig-height: 5.2

cond_plot <- df_conditions |>
  filter(conditions_category != "0 / Missing") |>
  mutate(conditions_category = factor(conditions_category, levels = conditions_category))

ggplot(cond_plot, aes(x = conditions_category, y = termination_rate)) +
  geom_col(fill = "#42616a", width = 0.55) +
  geom_text(
    aes(label = paste0(percent(termination_rate, accuracy = 0.1), "\n(", comma(terminated_studies), " trials)")),
    vjust = -0.25,
    color = "#183642",
    size = 3.4,
    lineheight = 0.9
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.085), expand = c(0, 0)) +
  labs(
    x = "Number of Conditions Listed in Registry",
    y = "Termination Rate",
    subtitle = "Single-condition and highly multi-morbid protocols (5+ conditions) each reach 5.8%–6.0% termination rates."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "#183642", size = 10),
    axis.title = element_text(color = "#183642", face = "bold"),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```

## 6. Study Duration: The Attrition Timeline

How does elapsed study timeline correlate with termination? The chart below tracks termination rates across study duration brackets (from short studies under 6 months to multi-year longitudinal trials):

```{r}
#| label: fig-durations
#| fig-cap: "Termination rate by total study duration bracket"
#| fig-width: 12.5
#| fig-height: 5.6

dur_plot <- df_durations |>
  mutate(duration_bracket = factor(duration_bracket, levels = rev(duration_bracket)))

ggplot(dur_plot, aes(x = termination_rate, y = duration_bracket)) +
  geom_col(fill = "#b8d8d8", color = "#183642", width = 0.65) +
  geom_text(
    aes(label = paste0(percent(termination_rate, accuracy = 0.1), " (", comma(terminated_studies), " trials)")),
    hjust = -0.06,
    color = "#183642",
    size = 3.3,
    fontface = "bold"
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.09), expand = expansion(mult = c(0, 0.16))) +
  labs(
    x = "Termination Rate",
    y = NULL,
    subtitle = "Trials lasting 2 to 5 years experience the highest termination rate (6.8%), facing chronic accrual fatigue and sponsor shifts."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", color = "#183642", size = 9.5),
    axis.text.x = element_text(color = "#42616a"),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```

::: {.callout-note}
### The 2–5 Year Vulnerability Window
Short studies (< 6 months) resolve quickly before major institutional or competitive shifts occur (3.4% failure). Studies extending into years 2 through 5 endure prolonged exposure to funding renewals, principal investigator relocations, and patient dropouts.
:::

## 7. Protocol Complexity Score: Compounding Structural Risk

To capture the holistic operational burden of a clinical trial, we synthesized a **Composite Study Complexity Score (0 to 5 points)** incorporating:
- Multi-arm design ($\ge 3$ arms or $\ge 3$ interventions)
- Placebo control reporting
- Blinded masking (double, triple, or quadruple)
- Multi-site geographic footprint ($\ge 6$ facilities)
- Multi-endpoint burden ($\ge 3$ primary outcomes)

Stratifying 602,514 studies by complexity score reveals a stark, linear escalation in failure risk:

```{r}
#| label: fig-complexity
#| fig-cap: "Trial termination rate across composite protocol complexity tiers"
#| fig-width: 12.5
#| fig-height: 5.6

comp_plot <- df_complexity |>
  mutate(complexity_tier = factor(complexity_tier, levels = complexity_tier))

ggplot(comp_plot, aes(x = complexity_tier, y = termination_rate)) +
  geom_col(fill = "#e76f51", width = 0.55) +
  geom_text(
    aes(label = paste0(percent(termination_rate, accuracy = 0.1), "\n(", comma(terminated_studies), " / ", comma(total_studies), ")")),
    vjust = -0.25,
    color = "#183642",
    size = 3.4,
    lineheight = 0.9
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.12), expand = c(0, 0)) +
  labs(
    x = "Protocol Complexity Tier",
    y = "Termination Rate",
    subtitle = "Termination risk nearly doubles from Low Complexity (5.2%) to Very High Complexity (9.3%)."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "#183642", size = 10),
    axis.title = element_text(color = "#183642", face = "bold"),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```

## 8. Sponsor Type & Site Network Footprint

### A. Lead Sponsor Category

```{r}
#| label: fig-sponsor
#| fig-cap: "Trial termination rate across lead sponsor categories"
#| fig-width: 12.5
#| fig-height: 5.6

sponsor_plot <- sponsor_summary |>
  filter(total_studies >= 1000) |>
  arrange(desc(termination_rate)) |>
  mutate(sponsor_label = factor(sponsor_label, levels = rev(sponsor_label)))

ggplot(sponsor_plot, aes(x = termination_rate, y = sponsor_label)) +
  geom_col(fill = "#42616a", width = 0.65) +
  geom_text(
    aes(label = paste0(percent(termination_rate, accuracy = 0.1), " (", comma(terminated_studies), " trials)")),
    hjust = -0.08,
    color = "#183642",
    size = 3.3
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.12), expand = expansion(mult = c(0, 0.15))) +
  labs(
    x = "Termination Rate",
    y = NULL,
    subtitle = "Industry (8.9%) and NIH-sponsored studies (8.5%) exhibit nearly double the termination rate of academic networks (4.8%)."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", color = "#183642", size = 9.5),
    axis.text.x = element_text(color = "#42616a"),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```

### B. Clinical Site Footprint

```{r}
#| label: fig-sites
#| fig-cap: "Termination rate across clinical site footprint categories"
#| fig-width: 12.5
#| fig-height: 5.4

site_plot <- site_summary |>
  filter(site_category != "0 sites / Not reported") |>
  mutate(site_category = factor(site_category, levels = c(
    "Single-site (1)",
    "Small multi-site (2–5)",
    "Medium multi-site (6–20)",
    "Large multi-site (21–50)",
    "Global mega-network (51+)"
  )))

ggplot(site_plot, aes(x = site_category, y = termination_rate)) +
  geom_col(fill = "#b8d8d8", color = "#183642", width = 0.6) +
  geom_text(
    aes(label = percent(termination_rate, accuracy = 0.1)),
    vjust = -0.5,
    color = "#183642",
    fontface = "bold",
    size = 3.6
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.13), expand = c(0, 0)) +
  labs(
    x = "Operational Footprint (Number of Clinical Sites)",
    y = "Termination Rate",
    subtitle = "Moving from a single site (5.0%) to multi-site networks (10.0%–10.3%) doubles termination risk."
  ) +
  theme_minimal(base_family = "DM Sans") +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "#183642", size = 9.5),
    axis.title = element_text(color = "#183642", face = "bold"),
    plot.subtitle = element_text(color = "#42616a", size = 9.5),
    plot.background = element_rect(fill = "#f7f4ed", color = NA),
    panel.background = element_rect(fill = "#f7f4ed", color = NA)
  )
```